Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-27 analysis/paper/paper.Rmd
remove figure 4 and 6, edit the relevant text

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by specialists attached to elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek201's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

@@

The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated. Reliable pottery outlines were captured by 13 harmonics that gather 99% of the total harmonic power. The average shape of vessels from each phase was visualized by computing the mean of the standardized Fourier coefficients within each phase group (Figure \@ref(fig:EFA-analysis)). Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods.

```{r EFA-analysis, fig.cap= "Mean pottery shapes of three phases"}
library(Momocs)
library(cowplot)
jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))

# default is 13 harmanics that gather 99% of the harmonic power 
op <- efourier(jpgs_imported_coo, norm = FALSE, 13)

@@

`Post-European` <- mshp$poe %T>% coo_plot(border="blue")
`Pre-European` <-  mshp$pre %T>% coo_plot(border="red", plot.new=FALSE)
`Chinese` <-       mshp$chi %T>% coo_plot(border="green", plot.new=FALSE)
legend("topright", 
       lwd=1,
       col=c("red", "blue", "green"), 
       legend=c( "Pre-European", "Post-European", "Chinese")) 
#? How to put the legend out of the plot, tried change the margin using par(), but doesn't work
```

@@

The elliptic Fourier coefficients were examined by Principal Components Analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variance. Figure \@ref(fig:contribution-PCs) shows the shape variation associated with the first three principal components. PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped. PC2 relates to the neck and mouth constriction, from narrow to wide. PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from curved to straight shape. The first two components account for most of the variance in relation to three phases (Figure \@ref(fig:first-two-PCs-byplot)). The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods. However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes. In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.

```{r contribution-PC-hide, results="hide", include= FALSE}
# the contributions of each PC to the shapes
pc_contrib <-  PCcontrib(op.p, nax = 1:3) 
```

```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes. The rows from the top to the bottom represent the first to third principal component respectively. Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  theme_minimal(base_size = 6)

ggsave(here::here("analysis/figures/pca-contrib.png"))

# PC1 is height of vessel
# PC2 is neck constriction
# PC3 is the degree of flare of neck
```

@@

To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \ref(@fig:first-two-PCs-byplot)), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases. Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$). Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other. There is little difference between vessel shapes in the Post-European and Chinese contact periods.

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images. We thank Dr. Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab. We thank the Quaternary Research Center funding for supporting the organic residue analysis in this project. We thank Dr. Julian Sachs in the Department of Oceanography, University of Washington for his supports and providing his lab for us to conduct geochemical analysis of potsherds. We thank Dr. Matthew Wolhowe for his help in developing protocols for lipid extraction and his assistance with the GC-MS, GC-FID, and GC-C-IRMS analyses. This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.

>>>>

EFA is suitable for shapes lacking representative landmarks or where curves contain the most meaningful variation. Applications in archaeology include human remains and zooarchaeology to understand  biological variation, and more recently on stone artifacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Ceramics is another archaeological materials that outline approaches can provide new insights into shape variation to answer anthropological questions related to ceramic specialization. Using semi-landmarks method, @Topi2018 examine 89 photographs of globular jars from the Casas Grandes of northwest Mexico. Their results suggest that some ceramic types from the Medio period (AD 1200-1450), were made by specialists based on highly standardized shapes. Among those highly standardized jars, some were made by specialists attached to elites indicating the presence of social inequality, while others were made by independent specialists according to their spatial distribution patterns. Similarity, @Selden2019 examined 28 Caddo bottle excavated at several sites (AD 500-1700) in northwest Louisiana using semi-landmarks method. The results indicate a significant integration between the different parts of the bottles including rim, neck, body, and base that vary in a coordinated manner. Also, two discrete base and body shapes represented northern and southern differences were identified. For the EFA method, @Wilczek2014 evaluate the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results show the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports that claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014's EFA results help us understand the level of production standardization over time across the region. Those examples show that outline approaches can distinguish variation in ceramic shapes at high resolutions to understand ceramic production.

@@

The elliptic Fourier coefficients of `r total_no` pottery from three phases were calculated. Reliable pottery outlines were captured by 13 harmonics that gather 99% of the total harmonic power. Figure \@ref(fig:thin-plate-splines) shows the shape changes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods. The average shape of vessels from each phase was visualized and it shows more obvious differences between post-European and Chinese periods.

# Mean pottery shapes of three phases
library(Momocs)
library(cowplot)
jpgs_imported_coo <- readr::read_rds(here::here("analysis/data/derived_data/jpgs_imported_coo.rds"))

# default is 13 harmonics that gather 99% of the harmonic power
op <- efourier(jpgs_imported_coo, norm = FALSE, 13)

@@

`Post-European` <- mshp$poe 
`Pre-European` <-  mshp$pre 
`Chinese` <-       mshp$chi

@@

The elliptic Fourier coefficients were examined by Principal Components Analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variance. According to Figure \@ref(fig:first-two-PCs-byplot), PC1 captures the height of the vessels, from tall to short, and the roundness of the body from round to oval-shaped. PC2 relates to the neck and mouth constriction, from narrow to wide. PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of the flare of the neck, from curved to straight shape. The first two components account for most of the variance in relation to three phases. The results reflect a large overlap in shapes from three occupations phases, especially for shapes in the pre-European and post-European periods. However, the spread of shape distribution indicates a wider variation in shapes in the pre-European and post-European periods compared to those in the Chinese period along both PC1 and PC2 axes. In other words, we find a decrease in shape variance in the Chinese period that is evident in the shorter height and narrower mouth of vessels used in that period.

@@

To test for differences in the distributions of shape variables indicated by the PC scores shown in Figure \@ref(fig:first-two-PCs-byplot), we used a multivariate analysis of variance (MANOVA) test to compare pairwise combinations across our three occupation phases. Table \@ref(tab:MANOVA-table) shows significant differences in shape between Pre-European and Post-European phases (p = $`r pval_post_pre`$), and Pre-European and Chinese phases (p = $`r pval_chi_pre`$). Although there is considerable overlap of shape variables between the Pre-European and Post-European phases, their PC scores are significantly different from each other. There is little difference between vessel shapes in the Post-European and Chinese contact periods.

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study and providing the shape images. We thank Dr. Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance of petrographic analysis at his lab. We thank the Quaternary Research Center funding for supporting the organic residue analysis in this project. We thank Dr. Julian Sachs in the Department of Oceanography, University of Washington for his supports and providing his lab for us to conduct organic geochemistry analysis of potsherds. We thank Dr. Matthew Wolhowe for his help in developing protocols for lipid extraction and his assistance with the GC-MS, GC-FID, and GC-C-IRMS analyses. This research used statistical consulting resources provided by the Center for Statistics and the Social Sciences, University of Washington.
